### 2010-04-15 thesis/SensorMLOnSenseTile/trunk/thesis/SensorMLonSenseTileThesis.tex
latest

\begin{abstract}
This project evaluates the Sensor Model Language(SensorML) and Sensor Observation Service(SOS) specifications. These specifications are intended for use in the development of open sensor systems. The UCD CASL SenseTile sensor system is used as a case study for this evaluation. A SensorML model of SenseTile is developed. A SenseTile Web Service is also developed based on the SenseTile SensorML model and the SOS specification. This Web Service provides access to SenseTile sensor metadata and sensor observations. The project finally examines formalizing SensorML using a formal specification language called Business Object Notation.

@@

The SenseTile Web Service developed in this project is based on the concepts described in the Sensor Observation Service (SOS) specification. The SOS Data Consumer and SOS Data Provider entities will referred to a DataConsumer and DataProvider respectively in this thesis when describing the SenseTile Web Service.

@@

\section{BON} \label{BONsec}
Business Object Notation (BON) is a notation and a method for analysis and design of Object-Oriented (OO) systems. It supports a seamless transition from the design to the source code as it based on the same Object Oriented (OO) concepts supported by most OO languages. It is reversible as changes in the source code can be easily mapped back into the design and analysis models. It supports design by contract\cite{Meyerref} where a function will guarantee a postcondition if a specified precondition is fulfilled by the caller of the function. BON support for design by contract allows for a formal description of a system to be developed. For a detailed description of BON see \cite{BONref}.
\chapter{Design}
\section{Overview}
This chapter describes the design of the SenseTile Web Service. Rather than define any new sensor models or XML formats for SenseTile it preferable to use the existing solutions if good enough for the task at hand. The OGC SWE framework as described in section\ref{SWESec} seems to cover all that is needed for modeling sensors and observations. Central to the design is a SensorML model of the SenseTile System.

@@

Of the SenseTile sensors only the Thermistor is modeled along with the  CelsiusConv Component that converts the digital number received from the Thermistor Component to a Celcius value. However the same approach can applied to the other senors though for some components no conversion process may be needed.

@@

The components and ComponentList elements contain a list of component elements. These component elements have a name attribute to reference the component within the SenseTile System. For example in this case the Sensor Board System is called sensorBoardSystem. The component can be a SensorML process or a processes chain.

@@

The SenseTile Web Service is based on the Sensor Observation Service (SOS) as described previously. The two main entities in the SOS Specification, the Sensor Data Provider and the SOS,  are used to structure the SenseTile Web Service. These entities are implemented as separate software components that together provide the service. The Sensor Data Provider component will be referred to as the DataProvider in this design.

@@

\subsection{SenseTile Web Service Scenarios}
This section contains UML sequence diagrams to show how the DataProducer and SOS interact to provide the sensor obsevatons and sensor descriptions to the DataConsumer client. There are three main scenarios supported by the SenseTile Web Service and they are described in the following sections.

@@

\newpage
\subsection{DataProducer}\label{DataProducerHigh}
The DataProducer's role, as described previously, is to read sensor data from the Sensor Board and generate O\&M observations. These observations are then sent to the SOS instance. The BON static diagram \ref{fig:bon_static_diagam_producer.png}
 shows the main clusters and classes of the DataProvider. BON notatation. Not all the classes are described.
\begin{figure}[h]
\centering
\scalebox{0.65}{\includegraphics{bon_static_diagam_producer.png}}
\caption{DataProducer BON Static Diagram}\label{fig:bon_static_diagam_producer.png}
\end{figure}

@@

\subsubsection {2. Process Sensor Data}
The PacketInputStream will request the SenseTileSystem to execute
the VastSMLEngine engine to process the sensor data. The VastSMLEngine
executes the process chains described in the SenseTile SensorML model.

@@

When the VastSMLEngine Engine is finished processing
the generated obsevations generated these
need to be sent to the SOS. An O\&M Onservation
XML string containing the sensor data is sent to
the SOS instance.

@@

The SOS component of the SenseTile Web Service provides a simplified OGC SOS Web interface to access the senosr observations from the DataProducer. The below is a BON static diagram showing the important clusters and classes in the SOS design.

 \begin{figure}[h!]
\centering
\scalebox{0.65}{\includegraphics{bon_static_diagam_provider.png}}
\caption{SOS BON Static Diagram}\label{fig:bon_static_diagam_provider.png}
\end{figure}
The SOS is the main cluster and the Service cluster contains the classes that provide the Web Service interface.

The DataProvider maintains a list of observation offerings provided by the SOS using the ObservationOffering class. When a new sensor type is registered a new ObservationOffering is created. If a DataProvider registers and provides an offering that already exists then just a new Sensor instance is added to the existing ObservationOffering.

The Sensor class contains a list of Observations that hold the observations from the DataProvider. The last 50 observations are maintained.

The ObservationService class provides the interface for the DataProducer to update with observations.

The SenseTile Service provides Web Service interface to be used by clients to access the sensor observations.

The SOS specification provides a very rich filtering feature that is too complex for this prototype. The operations are developed in a way to allow the sensor metadata and observations to be retrieved for a specific sensor board. The latest observation or all the stored observations can be retrieved.

\chapter{Implementation}
\section{Overview}
The SenseTile Web Service is written in Java. As described in the high level design there are two components, the DataProducer and the SOS, that are developed to provide this service. The two components are run in separate processes. The communication between the DataProducer and the SOS uses Java RMI. RMI is used due to the SOS and DataProducer
processes are running in the SenseTile Network and communicate over a LAN. This communication protocol has lower overhead than a Web Service so efficient communication between the DataProducer and SOS is achieved. Though it would not be hard to change to use a Web Service as a Java Interface is used to hide the communication mechanism from the main application code. Clients, SOS DataConsumers, access the SOS using a SOAP based Web Service interface.

@@

\section{SOS}
The SOS is the provider of the Web Service interface and is built on Apache Axis2 \footnote{http://ws.apache.org/axis2}. 
Axis2 is described as a "Web Services / SOAP / WSDL engine"  with the Apache Axis2/Java implementation used in this project. The Spring framework\footnote{http://www.springsource.org/} was also used. Using these libraries a light Web Service is built to run on the reasonably powerful SenseTile Processor Unit. Axis2 is run standalone here though it can be on with any Servlet engine if a more powerful Web Service is needed.

Developing a Web Service is straight forward with Axis2 and it supports several approaches. The supported POJOs (Plain Old Java Objects) approach was used for the SenseTile Web Service allowing the code to be reused on a different framework if AXIS2 did not work well. Axis2 supports the The Spring framework which is used in the SenseTile WebService for it's dependency injection functionality. A SprinitInit class is needed for Axis2 to load up the Spring framework.

Axis2 provides a tool that automatically generates WSDL from the running Web Service. Another tool allows client side code to be generated from the WSDL easing the development of the Web Service clients.

@@

\chapter{BON to SensorML mapping}
\section{Overview}
In this chapter a BON(see section\ref{BONsec}) to SensorML( see section \ref{SMLsection}) mapping is described. One motivation for this mapping is take advantage of formal software development processes described in \cite{Kiniryref}. This is a verification based process that refines BON to JML and then to Java. Robust software can thus be developed. Another motivation is that it XML is verbose and a more compact description of a SensorML model as facilitated by BON is naturally preferred. Lastly SensorML is complex and it is difficult to develop correct SenorML models based only on using the SensorML XML Schema definition. How a BON specification of SensorML can help here is described later in the chapter.

\section{SensorML BON Specification}
The BON specification of SensorML described is very close conceptually to the SensorML model shown in section \ref{SMLsection}.  The BON SensorML model is shown in figure \ref{fig:bonSensoMLModel}. 
\begin{figure}[h]
\centering
\scalebox{0.65}{\includegraphics{bonSensoMLModel.png}}
\caption{BON SensorML Model}\label{fig:bonSensoMLModel}
\end{figure}
The inheritance heirarhcy is changed slightly with System inheriting from ProcessChain and Component inheriting from ProcessModel. A new abstraction called Port is added. Port contains all the properties of SensorML input, output and parameter as they all have the same properties. A PortType class that ensures legal port types of Input, Output and Parameter is also added to the SensorML BON model.

A BON specification SWE Common data types Count and Quantity are also described along with needed dependent classes. The abstract class DataComponent is the base type for these. There are a many other types but only Count and Quantity are developed as they were the ones used in the SenseTile SensorML model.

The BON specification developed in this project does not address the issue of the SensorML metadata and the MetaDataGroup is not specified. This information is for descriptive purposes only and is not used for processing sensor data. BON adds no value here. A different way to add this information to SensorML sensor descriptions is suggested to be used.
AssociationAttributeGroup
SensorML
Member
The complete BON SensorML specification listing can be found in appendix \ref{appenB}.

\section{SensorML BON to XML Mapping}
SensorML is essential XML defined by XML Schema with its main purpose being interopeability. Therefore a mapping SensorML BON specification to the SensorML XML is needed. The BON types map closely to SensorML complex types and so a relatively simple scheme is needed.
Effective BON classes map to XML elements. Features in the class map either to attributes of the XML element or a contained XML element. Whether a feature is an attribute is indicated in the class as a comment using @SensorML to make this comment as a mapping instruction. Otherwise an XML element is assumed. The type names of the BON classes and features are the XML element or attribute names. Values. Case of names.

BON SEQUENCE types need a special rule for mapping to SensorML.

Connections rule.

@@

The source and destination reference strings need to be parsed to find out the inputs and outputs to connect. This is very error prone and the ability to check that connections are correct to a least some degree is desirable. A BON specification can enforce some level of correctness in setting up the connections using preconditions (ensure). The source reference must be either an input of the process chain itself or an output of one of the contained components. In the BON ProcessChain class a feature AddLink checks that Links are legal before adding the Link to the connections feature. This is shown in the BON listing below.

\begin{lstlisting}[basicstyle=\scriptsize,showspaces=false,showstringspaces=false, caption=BON ProcessChaing]  
effective class ProcessChain
     indexing
     about:        "Process formed by chaining sub-processes.";
     title:        "ProcessChain.";
     author:       "Ciaran Palmer.";
     copyright:    "none.";
     organisation: "School of Computer Science and Informatics, UCD.";
     date:         "2010/04/01.";
     version:      "Revision: 1.00.";

@@

\section{SenseTile Thermistor BON SensorML Specification }
Thermistor can be generated by the following BON class.

@@

A BON specification was developed for the core SensorML types and the SWE Common Data types used in the SenseTile SensorML model. The BON specification is included in appendix \ref{appenB} A BON to SensorML XML encoding mapping is also developed and is included in appendix \ref{appenC}.

@@

\subsubsection{SensorML}
SensorML provides a simple functional model of sensors in terms of an intuitive process concept. A process transforms sensor inputs to outputs as modeled by the SensorML AbstractProcess type. SensorML provides a process chaining capability to break up complex processing into smaller more manageable steps. There is also a rich set of data types to describe sensor data. Due to the simple model and the rich set of data types is is hard to imagine any sensor or sensor system that could not be described by SensorML. It is not intended to describle details of the sensor hardware. It is also not intended that SensorML provides the framework for encoding the actual sensor observations. This is provided by the SWE O\&M specification.

@@

SensorML can model a simple sensor or complex sensor systems. The SensorML specification defines a simple Detector model that limits the process to one input and one output. The Detector process outputs a real value such as a Celsius temperature value. eeicpr - tutorial ref. The SenseTile model uses a different approach as the unconverted data can also be retrieved from the SenseTile Web Service. There are good examples eecipr\footnote{} and tutorials available to help develop SensorML models.

The SenseTile model developed in this project is relatively simple and did not use the full richness of SensorML. It does not, for example, use the SensorML support for accurate and reliable processing data. It did examine a simple process chain which highlighted some of the difficulties of process chaining described in chapter x.

@@

The VastSensorMLEngine seems to be the reference implementation of a SensorML processing framework. There was little or no documentation that I could find on this and examples of it's use needed some research. It would be better if a working example could be provided with the framework and also some basic documentation. The VastSensorMLEngine code needed to be slightly modified to parse the SenseTile model files though they passed a validation check\footnote{http://vast.uah.edu/SensorMLforms/validate.jsp} provided by the Vast group. After these updates allowed the parsing to complete the code was indeed very stable and never crashed except when there were mistakes in the SenseTile SensorML XML files.

@@

\subsubsection{SensorML BON Specification}
The BON SensorML developed in this project just covers the core of the SenorML types. However it has shown how some of the problems of SensorML can be addressed with a formal language such as BON. A way to check that connections in SensorML processchains are correct is described using BON assertions. The type safety and assertions provided by a BON SensorML specification allows a better way to develop SensoML based sensor sysstems than just using the SensorML XML encoding.

A more complete BON SensorML description is needed than provided in this project and there are still many issues that need to be addressed. For example how to represent XLINKs in the SensorML BON specification or is there a need to. Based on the SensorML BON specification developed in this project an inline approach would be necessary rather than reference external XML files general used in the examples. For a complete SensorML description the SensorML metadata needs also to be included in an SensorML description. How to do this in not covered in this project.

@@

Streaming the data. See SWE Architecture document for a description of one approach.

Comparing a Restful approach to retrieving sensor observations rather than the SOS approach could be investigated. It may prove to be a simpler way to retrieve the sensor observations.

\newpage
\begin{thebibliography}{99}
\bibitem{OSWAref} Chu, Kobialka,  Durnota, and  Buyya, Open Sensor Web Architecture: Core Services
\bibitem{SMLref}Open Geospatial Consortium Inc., OpenGIS Sensor Model Language (SensorML) Implementation Specification, 2007
\bibitem{SOSref}Open Geospatial Consortium Inc.,  Sensor Observation Service, 2007
\bibitem{Kiniryref}Kim Waldén and Jean-Marc Nerson , "Seamless Object-Oriented Software Architecture", 1995
\bibitem{OMref}Open Geospatial Consortium Inc., Observations and Measurements – Part 1 - Observation schema, 2007
\bibitem{SWEArchref}Open Geospatial Consortium Inc., OGC Sensor Web Enablement Architecture, 2008
\bibitem{SMLTutorialref}Open Geospatial Consortium Inc., Using SensorML to describe a
Complete Weather Station , 2006
\bibitem{ProcessTutorialref}Open Geospatial Consortium Inc.,
\bibitem{OGCcatref}Open Geospatial Consortium Inc., OpenGIS® Catalogue Services Specification, 2007
\bibitem{Meyerref}Bertrand Meyer., "Object-Oriented Software Construction", 1997
\bibitem{Sensorsref}Bertrand Meyer., ", 1997

>>>>

\begin{abstract}
This project evaluates the Sensor Model Language(SensorML) and Sensor Observation Service(SOS) specifications. These specifications are intended for use in the development of open sensor systems. The UCD CASL SenseTile Sensor System is used as a case study for this evaluation. A SensorML model of SenseTile is developed. A SenseTile Web Service is also developed based on the SenseTile SensorML model and the SOS specification. This Web Service provides access to SenseTile sensor metadata and sensor observations. The project finally examines formalizing SensorML types using Business Object Notation.

@@

The SenseTile Web Service developed in this project is based on the concepts described in the Sensor Observation Service (SOS) specification. The SOS Data Consumer and SOS Data Provider entities are referred to a DataConsumer and DataProvider respectively in this thesis when describing the SenseTile Web Service.

@@

\section{BON} \label{BONsec}
Business Object Notation (BON) is a notation and a method for analysis and design of Object-Oriented (OO) systems. It supports a seamless transition from the design to the source code as it based on the same Object Oriented (OO) concepts supported by most OO languages. It is reversible as changes in the source code can be easily mapped back into the design and analysis models. It also supports design by contract\cite{Meyerref} where a function guarantees a postcondition if a specified precondition is fulfilled by the caller of the function. BON support for design by contract allows for a formal description of a system to be developed. For a detailed description of BON see \cite{BONref}.
\chapter{Design}
\section{Overview}
This chapter describes the design of the SenseTile Web Service. Rather than define any new sensor models or XML formats for SenseTile it preferable to use the existing solutions if good enough for the task at hand. The OGC SWE framework as described in section\ref{SWESec} seems to cover all that is needed for modeling sensors and observations. Central to the design is a SensorML model of the SenseTile System.

@@

Of the SenseTile sensors only the Thermistor is modeled along with the  CelsiusConv Component that converts the digital number received from the Thermistor Component to a Celsius value. However the same approach can applied to the other senors though for some components no conversion process may be needed.

@@

The components and ComponentList elements contain a list of component elements. These component elements have a name attribute to reference the component within the SenseTile System. For example in this case the Sensor Board System is called sensorBoardSystem. The component can be a SensorML process or a process chain.

@@

The SenseTile Web Service is based on the Sensor Observation Service (SOS) as described previously. The two main entities in the SOS Specification, the Sensor Data Provider and the SOS,  are used to structure the SenseTile Web Service. These entities are implemented as separate software components that together provide the service. The Sensor Data Provider component is referred to as the DataProvider in this design.

@@

\subsection{SenseTile Web Service Scenarios}
This section contains UML sequence diagrams to show how the DataProducer and SOS interact to provide the sensor observations and sensor descriptions to the DataConsumer client. There are three main scenarios supported by the SenseTile Web Service and they are described in the following sections.

@@

\newpage
\subsection{DataProducer}\label{DataProducerHigh}
The DataProducer's role, as described previously, is to read sensor data from the Sensor Board and generate O\&M observations. These observations are then sent to the SOS instance. The BON static diagram \ref{fig:bon_static_diagam_producer.png}
 shows the main clusters and classes of the DataProvider. BON notatation. Not all the classes are described.
\begin{figure}[h]
\centering
\scalebox{0.65}{\includegraphics{bon_static_diagam_producer.png}}
\caption{DataProducer BON Class Diagram}\label{fig:bon_static_diagam_producer.png}
\end{figure}

@@

\subsubsection {2. Process Sensor Data}
The PacketInputStream requests the SenseTileSystem to execute
the VastSMLEngine engine to process the sensor data. The VastSMLEngine
executes the process chains described in the SenseTile SensorML model.

@@

When the VastSMLEngine Engine is finished processing
the generated observations generated these
need to be sent to the SOS. An O\&M Observation
XML string containing the sensor data is sent to
the SOS instance.

@@

The SOS component of the SenseTile Web Service provides a simplified OGC SOS Web interface to access the sensor observations from the DataProducer. The below is a BON static diagram showing the important clusters and classes in the SOS design.

 \begin{figure}[h!]
\centering
\scalebox{0.65}{\includegraphics{bon_static_diagam_provider.png}}
\caption{SOS BON Class Diagram}\label{fig:bon_static_diagam_provider.png}
\end{figure}

The DataProvider class maintains a list of observation offerings provided by the SOS using the ObservationOffering class. When a new sensor type is registered a new ObservationOffering is created. If a DataProvider registers and provides an offering that already exists then just a new Sensor instance is created and its identity is added to the appropriate ObservationOffering.

The DataProducerService class provides the interface for the DataProducer to register senors and update the SOS with observations. These sensor observations are stored in  the Sensor class in a list of Observations classes. The SenseTileService provides the Web Service interface used by clients to access the sensor observations.

The SOS specification provides a very rich filtering feature in the GetObservation operation. However SenseTile Web Service only implements the functionality to get a sensor observation using a sensors identity.

\chapter{Implementation}
\section{Overview}
The SenseTile Web Service is written in Java. As described in the design chapter there are two components, the DataProducer and the SOS, that are developed to provide this service. The two components are run in separate processes. The communication between the DataProducer and the SOS uses Java RMI\footnote{http://java.sun.com/javase/technologies/core/basic/rmi/index.jsp}. RMI is used due to the SOS and DataProducer
processes are running in the SenseTile Network and communicate over a LAN. This communication has lower overhead than using HTTP so efficient communication between the DataProducer and SOS is achieved. It is not be hard to change to use a a different transport layer as a Java Interface is used to hide the communication mechanism from the main application code. Clients, SOS DataConsumers, access the SOS using a SOAP\footnote{http://www.w3.org/TR/soap/} based Web Service interface.

@@

\section{SOS}
The SOS is the provider of the Web Service interface and is built on Apache Axis2 \footnote{http://ws.apache.org/axis2}. 
Axis2 is described as a "Web Services / SOAP / WSDL engine"  with the Apache Axis2/Java implementation used in this project. The Spring framework\footnote{http://www.springsource.org/} was also used. Using these technologies a light Web Service is built to run on the reasonably powerful SenseTile Processor Unit. Axis2 is run standalone here though it can run on any Servlet engine if a more powerful Web Service is needed.

Developing a Web Service is straight forward with Axis2 and it supports several approaches to developing Web applications. The supported POJOs (Plain Old Java Objects) approach was used for the SenseTile Web Service allowing the code to be reused on a different framework if AXIS2 did not work well. Axis2 supports the Spring framework which is used in the SenseTile WebService for it's dependency injection functionality. A SprinitInit class is needed for Axis2 to load up the Spring framework.

Axis2 provides a tool that automatically generates WSDL\footnote{http://www.w3.org/TR/wsdl} from the running Web Service. Another tool allows client side code to be generated from the WSDL allowing easier development of the Web Service client.

@@

\chapter{SensorML BON Specifications}
\section{Overview}
In this chapter a BON(see section\ref{BONsec}) SensorML( see section \ref{SMLsection}) specification is described. One motivation for this is take advantage of formal software development processes described in \cite{Kiniryref}. This is a verification based process that refines BON to JML and then to Java. Robust software can thus be developed. Another motivation is that it XML is verbose and a more compact description of a SensorML model as facilitated by BON is preferred. Lastly SensorML is complex and it is difficult to develop correct SensorML models based only on using the SensorML XML Schema definition. A BON specification of SensorML can help here and an example of this is described later in the chapter.

\section{SensorML BON Specification}
The BON specification of SensorML described is very close conceptually to the SensorML model shown in section \ref{SMLsection}.  The SensorML BON Class Diagram is shown in figure \ref{fig:bonSensoMLModel}. 
\begin{figure}[h]
\centering
\scalebox{0.65}{\includegraphics{bonSensoMLModel.png}}
\caption{SensorML BON Class Diagram}\label{fig:bonSensoMLModel}
\end{figure}
\newline
The inheritance hierarchy is changed with System inheriting from ProcessChain and Component inheriting from ProcessModel. A new abstraction called Port is added. Port contains all the properties to describe SensorML process's input, output and parameter. A PortType class that ensures legal port types of Input, Output and Parameter is also added to the SensorML BON model.

A BON specification SWE Common data types Count and Quantity are also described along with needed dependent classes. The abstract class DataComponent is the base type for these. There are a many other types but only Count and Quantity are developed as they are the ones used in the SenseTile SensorML model.

The BON specification developed in this project does not address the issue of the SensorML metadata and the MetaDataGroup is not specified. This information is for descriptive purposes only and is not used for processing sensor data. BON adds no value here. A different way to add this information to SensorML sensor descriptions is suggested to be used.

eeicpr

AssociationAttributeGroup
SensorML
Member

The SensorML BON specifications developed in this project are be found in appendix \ref{appenB}.

@@

The source and destination reference strings need to be parsed to find out the inputs and outputs to connect. This is very error prone and the ability to check that connections are correct to a least some degree is desirable. A BON specification can enforce some level of correctness in setting up the connections using preconditions. The source reference must be either an input of the process chain itself or an output of one of the contained components. In the BON ProcessChain class a feature AddLink checks that Links are legal before adding the Link to the connections feature. This is shown in the BON ProcessChain listing below.

\begin{lstlisting}[basicstyle=\scriptsize,showspaces=false,showstringspaces=false, caption=BON ProcessChain]  
effective class ProcessChain
     indexing
     about:        "Process formed by chaining sub-processes.";
     title:        "ProcessChain.";
     author:       "Ciaran Palmer.";
     copyright:    "none.";
     organisation: "School of Computer Science and Informatics, UCD.";
     date:         "2010/04/01.";
     version:      "Revision: 1.00.";

@@

@@

A SensorML BON specification was developed for the core SensorML types and the SWE Common Data types used in the SenseTile SensorML model. This specification is included in appendix \ref{appenB}.

@@

\subsubsection{SensorML}
SensorML provides a simple functional model of sensors in terms of an intuitive process concept. A process transforms sensor inputs to outputs as modeled by the SensorML AbstractProcess type. SensorML provides a process chaining capability to break up complex processing into smaller more manageable steps. There is also a rich set of data types to describe sensor data. Due to the simple model and the rich set of data types is is hard to imagine any sensor or sensor system that could not be described by SensorML. It is not intended to describe details of the sensor hardware. It is also not intended that SensorML provides the framework for encoding the actual sensor observations. This is provided by the SWE O\&M specification.

@@

SensorML can model a simple sensor or complex sensor systems. The SensorML specification defines a simple Detector model that limits the process to one input and one output. The Detector process outputs a real value such as a Celsius temperature value. The SenseTile model uses a different approach as the unconverted data can also be retrieved from the SenseTile Web Service. There are good examples\footnote{http://www.ogcnetwork.net/SWE\_Projects} and tutorials \cite{SMLTutorialref} available to help develop SensorML models.

The SenseTile model developed in this project is relatively simple and did not use the full richness of SensorML. It does not, for example, use the SensorML support for accurate and reliable processing data. It did examine a simple process chain which highlighted some of the difficulties of process chaining described in chapter \ref{BONsec}..

@@

The VastSensorMLEngine seems to be the reference implementation of a SensorML processing framework. There was little or no documentation that I could find on this and examples of it's use needed some research. It would be good if a working example could be provided with the framework and also some basic documentation. The VastSensorMLEngine code needed to be slightly modified to parse the SenseTile model files though they passed a validation check\footnote{http://vast.uah.edu/SensorMLforms/validate.jsp} provided by the Vast group. After these updates allowed the parsing to complete the code was indeed very stable and never crashed except when there were mistakes in the SenseTile SensorML XML files.

@@

\subsubsection{SensorML BON Specification}
The BON SensorML developed in this project just covers the core of the SensorML types. However it has shown how some of the problems of SensorML can be addressed with a formal language such as BON. A way to check that connections in SensorML processchains are correct is described using BON assertions. The type safety and assertions provided by a BON SensorML specification allows a better way to develop SensorML based sensor sysstems than just using the SensorML XML encoding.

A more complete BON SensorML description is needed than provided in this project and there are still many issues that need to be addressed. For example how to represent XLinks in the SensorML BON specification or is there a need to. Based on the SensorML BON specification developed in this project an inline approach would be necessary rather than reference external XML files general used in the examples. For a complete SensorML description the SensorML metadata needs also to be included in an SensorML description. How to do this in not covered in this project.

@@

Comparing a RESTful\cite{Restref} approach to retrieving sensor observations rather than the SOS approach could be investigated. It may prove to be a simpler way to retrieve the sensor observations.

\newpage
\begin{thebibliography}{99}
\bibitem{OSWAref} Chu, Kobialka,  Durnota, and  Buyya, Open Sensor Web Architecture: Core Services
\bibitem{SMLref}Open Geospatial Consortium Inc., OpenGIS Sensor Model Language (SensorML) Implementation Specification, 2007
\bibitem{SOSref}Open Geospatial Consortium Inc.,  Sensor Observation Service, 2007
\bibitem{Kiniryref}Kim Waldén and Jean-Marc Nerson , "Seamless Object-Oriented Software Architecture", 1995
\bibitem{OMref}Open Geospatial Consortium Inc., Observations and Measurements – Part 1 - Observation schema, 2007
\bibitem{SWEArchref}Open Geospatial Consortium Inc., OGC Sensor Web Enablement Architecture, 2008
\bibitem{SMLTutorialref}Open Geospatial Consortium Inc., Using SensorML to describe a
Complete Weather Station , 2006
\bibitem{OGCcatref}Open Geospatial Consortium Inc., OpenGIS® Catalogue Services Specification, 2007
\bibitem{Meyerref}Bertrand Meyer., "Object-Oriented Software Construction", 1997
\bibitem{Restref}Fielding, Roy Thomas (2000), Architectural Styles and the Design of Network-based Software Architectures, Doctoral dissertation, University of California, Irvine
